Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-21 paper.tex
Created table for GA for feature selection
Please Mario check that what is said is correct. Also try to justify a
bit more the selection of parameters.

To calculate the average and standard deviations of these key presses, the delays
between a key-down and a key-up event of the left button clicks are used. In
addition to these averages and standard deviations of the delays between
keystrokes and mouse button presses, the average and standard deviations of the
number of total events contained in a digraph and a trigraph are calculated.
These features are proposed and explained by Epp, Lippold y Mandryk \cite{epp2011identifying}. Most
of the times, a digraph should contain four events, while a trigraph six.
However, sometimes an individual can start a digraph or a trigraph before ending
the previous one. These additional features represent these particular cases,
and could be meaningful for the estimation of a learner\'s affective states.
Regarding the mouse movements, the average and standard deviation of the
duration of each mouse movement, and the averages and standard deviations of the
movements in the axes X and Y are also calculated. Lastly, a final feature is
added to preprocessing of the data. The web tutorial recorded  and included in the feature vector how many attempts
a student required before successfully solving an assignment. The final feature vector
consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
\cite{epp2011identifying}.
%and are shown in Table \ref{tab_features}.

@@

As part of the data mining process a canonical genetic algorithm was used for feature
selection with the following parameters: population size 700, 30 generations with
tournament selection, tournament size was 25\% of population size. A uniform crossover
with 0.5 probability and mutation rate of 1/39, and a minimum feature size of 5.
In the end, the subset for the frustration classifier includes only 11 features, and 
the subset for the boredom classifier consists of only 13 features.

@@

Table \ref{tab_perf_flow} shows the performance for the additional classifiers
evaluated specifically for the Flow/engaged affective state. The goal was to
increase the performance achieved earlier. Again a ten-
fold cross-validation was conducted. In this case, instead of using an
evolutionary algorithm to optimize feature selection, it was used for 
optimizing a weight vector. 
The feature-weight vector assigned a weight to each o the features. The
parameters used for the evolutionary optimization algorithm are the following:
Population size = 10, generations = 30, selection = tournament, tournament size = 0.25 (fraction of the population),  initialize probability = 0.5,  mutation
probability = 1/number of attributes,  crossover probability = 0.5. Parameters
used for the Random Forest algorithm: number of trees = 10, 
criterion = gain ratio, maximal depth = 20, 
confidence = 0.25, minimal gain = 0.1, minimal leaf size = 2.  The
parameters used for are the J48 are the following:  confidence = 0.79, minimal
leaf size=2. For the kNN algorithm, the value k = 4 was used. 
For gradient boosted
trees: number of trees = 20,  maximal depth = 5, minimum rows = 10, 
minimum split 
improvement = 0.0, number of bins = 20, 
learning rate = 0.1, sample rate = 1.0.  Bayes
Kernel: laplace correction = True, estimation mode = greedy, 
minimum bandwidth = 0.1, number of kernels = 10.  Finally for deep learning: activation = Rectifier, hidden layer sizes = [50, 50], epochs = 10.

@@

%This
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text?
% Is 80% effective?
%  - done  Mario 
Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
\cite{blikstein2011using}; the parameters proposed by this author
include the size of the program and changes in the code. Other authors
also propose treating the use of certain keys as features e.g., the
delete key because is manly used for correcting sentences. Also a multi-modal approach is needed. Since all these need a completely new approach, a new methodology will be developed to
gather data and take measures.

>>>>

To calculate the average and standard deviations of these key presses, the delays
between a key-down and a key-up event of the left button clicks are used. In
addition to these averages and standard deviations of the delays between
keystrokes and mouse button presses, the average and standard deviations of the
number of total events contained in a digraph and a trigraph are calculated.
These features are proposed and explained by Epp, Lippold y Mandryk \cite{epp2011identifying}. Most
of the times, a digraph should contain four events, while a trigraph six.
However, sometimes an individual can start a digraph or a trigraph before ending
the previous one. These additional features represent these particular cases,
and could be meaningful for the estimation of a learner\'s affective states.
Regarding the mouse movements, the average and standard deviation of the
duration of each mouse movement, and the averages and standard deviations of the
movements in the axes X and Y are also calculated. Lastly, a final feature is
added to preprocessing of the data. The web tutorial recorded  and included in the feature vector how many attempts
a student required before successfully solving an assignment. The final feature vector
consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
\cite{epp2011identifying} and are shown in Table \ref{tab_features}.

@@

\begin{table}[h!tb]
\centering
\caption{Parameters used in the evolutionary algorithm for feature
  selection.}
\label{tab_ga}
    \begin{tabular}{ | l | l | }
      \hline
      Parameter & Value \\
      Population size & 700 \\
      Generations & 30 \\
      Selection & Tournament \\
      Tournament size & 25\% of population size \\
      Crossover & Uniform \\
      Crossover probability & 0.5 \\
      Mutation rate & $1 \over 39$ \\
      Minimum number of features & 5 \\
    \end{tabular}
  \end{table}

As part of the data mining process a canonical genetic algorithm was used for feature
selection with the parameters used shown in Table \ref{tab_ga}. These
parameters were set heuristically, and were the same for selecting the
features of all classification algorithms. % Mario check this - JJ
In the end, the subset for the frustration classifier includes only 11 features, and 
the subset for the boredom classifier consists of only 13 features.

@@

Table \ref{tab_perf_flow} shows the performance for the additional classifiers
evaluated specifically for the Flow/engaged affective state. The goal was to
increase the performance achieved earlier. Again a ten-
fold cross-validation was conducted. In this case, instead of using an
evolutionary algorithm to optimize feature selection, it was used for 
optimizing a weight vector. 
The feature-weight vector assigned a weight to each o the features. The
parameters used for the evolutionary optimization algorithm are the following:
Population size = 10, generations = 30, selection = tournament, tournament size = 0.25 (fraction of the population),  initialize probability = 0.5,  mutation
probability = 1/number of attributes,  crossover probability = 0.5. Parameters
used for the Random Forest algorithm: number of trees = 10, 
criterion = gain ratio, maximal depth = 20, 
confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2.  The
parameters used for are the J48 are the following:  confidence = 0.79, minimal
leaf size=2. For the kNN algorithm, the value k = 4 was used. 
For gradient boosted
trees: number of trees = 20,  maximal depth = 5, minimum rows = 10, 
minimum split 
improvement = 0.0, number of bins = 20, 
learning rate = 0.1, sample rate = 1.0.  Bayes
Kernel: laplace correction = True, estimation mode = greedy, 
minimum bandwidth = 0.1, number of kernels = 10.  Finally for deep learning: activation = Rectifier, hidden layer sizes = [50, 50], epochs = 10.

@@

%This
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text?
% Is 80% effective?
%  - done  Mario 
Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
\cite{blikstein2011using}; the parameters proposed by this author
include the size of the program and changes in the code. Other authors
also propose treating the use of certain keys as features e.g., the
delete key because it is mainly used for correcting sentences. Also a multi-modal approach is needed. Since all these need a completely new approach, a new methodology will be developed to
gather data and take measures.
